---
title: "R Compilation Analysis"
output: html_document
editor_options: 
  chunk_output_type: console
params:
  base_dir: "."
---

Run with

```
Rscript -e "rmarkdown::render('analysis.Rmd', output_file = 'analysis.html', params = list(base_dir = '.'))"
```

## Setup

```{r setup}
library(tidyverse)
library(texvars)
```

## Loading Data

We start by loading the compilation results from the `results.csv` files located in each package directory.

```{r load-data}
# Find all results.csv files
csv_files <- list.files(path = params$base_dir, pattern = "results.csv", recursive = TRUE, full.names = TRUE)

if (length(csv_files) == 0) {
  stop(paste("No 'results.csv' files found in", params$base_dir, "or its subdirectories."))
}

# Define column types based on the provided sample
col_spec <- readr::cols(
  name = readr::col_character(),
  symbol_name = readr::col_character(),
  binary_size = readr::col_double(),
  constants_size = readr::col_double(),
  compile_time_ms = readr::col_double(),
  bytecode_instructions = readr::col_double(),
  bytecode_compile_time_ms = readr::col_double(),
  c_compile_time_ms = readr::col_double(),
  native_compile_time_ms = readr::col_double(),
  object_file = readr::col_character(),
  constants_file = readr::col_character(),
  source_file = readr::col_character(),
  error = readr::col_character(),
  warning = readr::col_character()
)

all_results <- map_dfr(csv_files, ~{
  package_name <- basename(dirname(.x))
  tryCatch({
    readr::read_csv(.x, col_types = col_spec) %>%
      mutate(package = package_name)
  }, error = function(e) {
    warning(paste("Could not read or process", .x, ":", e$message))
    return(NULL)
  })
}, .progress = TRUE) %>% 
  mutate(across(c(object_file, constants_file, source_file), ~{
    # Normalize paths to contain only <last_dir>/<file>
    # This assumes the file is inside a package directory which is inside the base_dir structure
    if_else(is.na(.x), NA_character_, file.path(basename(dirname(.x)), basename(.x)))
  }))

if (nrow(all_results) == 0) {
  stop("All 'results.csv' files failed to load which is weird to say the least")
}
```

## Analyzing Lines of Code

Next, we analyze the lines of code in the generated `.c` files using `cloc`.

```{r loc}
# Check if cloc is installed
cloc_path <- Sys.which("cloc")
if (cloc_path == "") {
  if (file.exists("/opt/homebrew/bin/cloc")) {
    cloc_path <- "/opt/homebrew/bin/cloc"
  } else {
    stop("`cloc` is not installed. Please install it and ensure it's in your PATH.")
  }
}

# all known C files normalized to "package/file.c"
source_files_rel <- all_results %>%
  filter(!is.na(source_file)) %>%
  pull(source_file) %>%
  unique()

source_files_full <- file.path(params$base_dir, source_files_rel)

existing_files <- file.exists(source_files_full)
missing_files <- source_files_rel[!existing_files]

if (length(missing_files) > 0) {
  warning(paste("The following", length(missing_files), "source files were not found and will be skipped by cloc:\n", paste(head(missing_files, 10), collapse = "\n"), "\n..."))
}

files_to_cloc <- source_files_full[existing_files]

if (length(files_to_cloc) == 0) {
  stop("No valid C source files found to process with cloc.")
}

# create a temporary file to hold the list of files for cloc
c_files_list <- tempfile(fileext = ".txt")
writeLines(files_to_cloc, c_files_list)

# run cloc on the list of files to get a single CSV output
cloc_csv_output <- system(paste0(cloc_path, " --by-file --csv --list-file=", c_files_list), intern = TRUE)
cloc_data <- read_csv(
  I(cloc_csv_output[-1]), # cloc does not generate proper CSV header
  col_names = c(
    "language",
    "filename",
    "blank",
    "comment",
    "code"
  )
)

if (nrow(cloc_data) == 0) {
  stop("No C code files were processed by cloc")
}

loc <- cloc_data %>%
  filter(language == "C") %>%
  select(source_file = filename, c_code_lines = code) %>%
  mutate(source_file = file.path(basename(dirname(source_file)), basename(source_file)))
```

## Combined Analysis

```{r combined-analysis}
D <- left_join(all_results, loc, by = "source_file")
```

```{r}
classify_error <- function(x) {
  case_when(
    str_detect(x, "Received message larger than max") ~ "Message to large",
    str_detect(x, "Duplicate definition of symbol") ~ "Duplicate symbol",
    str_detect(x, "Cannot native compile function") ~ "C compile error",
    str_detect(x, "Empty bytecode for function") ~ "Empty bytecode",
    str_detect(x, "Cannot bytecode compile function") ~ "Bytecode compile error",
    is.na(x) ~ NA,
    TRUE ~ "Other"
  )
}

D <- mutate(D, error = classify_error(error))
```
## Summary Statistics

This section provides a summary of the compilation errors and descriptive statistics for the numeric variables in the dataset.

### Packages Summary

```{r}
count(D, package, sort = TRUE) %>% 
  knitr::kable(
    caption = "Number of compilation results per package.",
    col.names = c("Package", "Functions")
  )
```


### Compilation Summary

```{r error-summary, echo=FALSE, message=FALSE, warning=FALSE}
error_summary <- D %>%
  count(error, sort = TRUE, name = "count") %>%
  mutate(
    cumulative_count = cumsum(count),
    percentage = (count / sum(count)) * 100,
    cumulative_percentage = cumsum(percentage),
    error = replace_na(error, "No Error") 
  ) %>%
  mutate(
    count = scales::comma(count),
    cumulative_count = scales::comma(cumulative_count)
  )

knitr::kable(
  error_summary,
  caption = "Frequency and cumulative distribution of compilation errors.",
  col.names = c("Error Message", "Count", "Cumulative Count", "%", "Cumulative %"),
  digits = 2
)
```

### Numeric Summary

```{r numeric-summary, echo=FALSE, message=FALSE, warning=FALSE}
numeric_summary <- D %>%
  select_if(is.numeric) %>%
  summarise(across(everything(),
                   list(mean = mean, min = min, max = max, median = median, sd = sd),
                   na.rm = TRUE,
                   .names = "{.col}___{.fn}")) %>%
  pivot_longer(everything(),
               names_to = c("variable", "statistic"),
               names_sep = "___") %>%
  pivot_wider(names_from = statistic, values_from = value)

formatted_summary <- numeric_summary %>%
  select(-sd) %>% 
  pivot_longer(cols = -variable, names_to = "statistic", values_to = "value") %>%
  mutate(
    formatted_value = case_when(
      grepl("size", variable) ~ texvars::size_1000(value) %>% fmt(),
      grepl("_ms", variable) ~ texvars::duration(value/1000),
      grepl("instructions", variable) ~ fmt(value, digits=1),
      TRUE ~ fmt(value)
    )
  ) %>%
  select(-value) %>%
  pivot_wider(names_from = statistic, values_from = formatted_value) %>% 
  mutate(variable = case_match(
      variable,
      "binary_size" ~ "Binary size",
      "constants_size" ~ "Constants size",
      "compile_time_ms" ~ "Compile time",
      "bytecode_instructions" ~ "Bytecode instructions",
      "bytecode_compile_time_ms" ~ "Bytecode compile time",
      "c_compile_time_ms" ~ "C compile time",
      "native_compile_time_ms" ~ "Native compile time",
      "c_code_lines" ~ "C code size (lines)"
    ))

knitr::kable(
  formatted_summary,
  caption = "Summary statistics for numeric variables in the dataset."
)
```